---
title: "HW2"
author: "Isha Pargaonkar"
date: "2026-02-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Source and Citation

This is the Meteorite Landings dataset from NASA's Open Data Portal, which contains comprehensive information on all known meteorite landings documented by The Meteoritical Society. The dataset includes over 45,000 meteorite records spanning from 860 AD to 2013, with detailed information about location, mass, classification, and discovery circumstances.

Data Source: NASA Open Data Portal

URL: <https://data.nasa.gov/dataset/meteorite-landings>

Primary Source: This is the original dataset maintained by NASA

## Import Dataset

```{r}
install.packages("maps")
```

```{r import-data}
# Load required libraries
library(ggplot2)
library(maps)

# Read dataset
meteorites <- read.csv("Meteorite_Landings.csv", stringsAsFactors = FALSE)

# Display first 10 rows
head(meteorites, 10)
```

## Variables of Interest

For this analysis, I am interested in exploring **how different regions discover different types of meteorites over time**. The key variables used are:

-   **`name`**: Name identifier for the meteorite
-   **`reclat`**: Latitude coordinate (decimal degrees) where the meteorite was found
-   **`reclong`**: Longitude coordinate (decimal degrees) where the meteorite was found
-   **`year`**: Year the meteorite was discovered or observed falling
-   **`mass..g.`**: Mass of the meteorite in grams
-   **`fall`**: Classification indicating whether the meteorite was observed falling ("Fell") or discovered later ("Found")
-   **`recclass`**: Detailed classification of the meteorite (e.g., H5, Iron IVB, Pallasite, Achondrite) - **this is the key variable for determining meteorite type**

## Data Processing Steps

The following processing steps were performed to prepare the data for visualization:

```{r process-data}
# 1. Remove rows with missing coordinates, year, or classification
meteorites_clean <- meteorites[!is.na(meteorites$reclat) & 
                                !is.na(meteorites$reclong) & 
                                !is.na(meteorites$year) &
                                !is.na(meteorites$recclass), ]

# 2. Remove outliers (coordinates outside valid ranges)
meteorites_clean <- meteorites_clean[meteorites_clean$reclat >= -90 & 
                                     meteorites_clean$reclat <= 90 & 
                                     meteorites_clean$reclong >= -180 & 
                                     meteorites_clean$reclong <= 180, ]

# 3. Focus on discoveries from 1800 onwards for clearer temporal patterns
meteorites_clean <- meteorites_clean[meteorites_clean$year >= 1800 & 
                                     meteorites_clean$year <= 2013, ]

# 4. Create time period categories for comparison
meteorites_clean$time_period <- ifelse(meteorites_clean$year < 1950, 
                                       "1800-1949: Early Discoveries",
                                ifelse(meteorites_clean$year >= 1950 & meteorites_clean$year < 1980,
                                       "1950-1979: Mid-Century",
                                ifelse(meteorites_clean$year >= 1980 & meteorites_clean$year < 2000,
                                       "1980-1999: Modern Era",
                                       "2000-2013: Recent Discoveries")))

# 5. Convert to factor with ordered levels for proper plotting
meteorites_clean$time_period <- factor(meteorites_clean$time_period, 
                                       levels = c("1800-1949: Early Discoveries",
                                                  "1950-1979: Mid-Century", 
                                                  "1980-1999: Modern Era",
                                                  "2000-2013: Recent Discoveries"))

# 6. Create simplified meteorite type categories from detailed recclass
# Classify meteorites into major groups based on their composition
meteorites_clean$meteorite_type <- ifelse(grepl("^Iron", meteorites_clean$recclass) | 
                                          grepl("^Ir", meteorites_clean$recclass), 
                                          "Iron",
                                   ifelse(grepl("^Stone", meteorites_clean$recclass) |
                                          grepl("^H[0-9]", meteorites_clean$recclass) |
                                          grepl("^L[0-9]", meteorites_clean$recclass) |
                                          grepl("^LL[0-9]", meteorites_clean$recclass) |
                                          grepl("^E[0-9]", meteorites_clean$recclass) |
                                          grepl("Chondrite", meteorites_clean$recclass) |
                                          grepl("Achondrite", meteorites_clean$recclass) |
                                          grepl("^C[0-9]", meteorites_clean$recclass),
                                          "Stone",
                                   ifelse(grepl("Stony-iron", meteorites_clean$recclass) |
                                          grepl("Pallasite", meteorites_clean$recclass) |
                                          grepl("Mesosiderite", meteorites_clean$recclass),
                                          "Stony-Iron",
                                          "Other")))

meteorites_clean$meteorite_type <- factor(meteorites_clean$meteorite_type,
                                          levels = c("Stone", "Iron", "Stony-Iron", "Other"))

# 7. Create region categories for geographic analysis
meteorites_clean$region <- ifelse(meteorites_clean$reclat > -70 & meteorites_clean$reclat < -60,
                                  "Antarctica",
                           ifelse(meteorites_clean$reclat >= 15 & meteorites_clean$reclat <= 75 & 
                                  meteorites_clean$reclong >= -25 & meteorites_clean$reclong <= 60,
                                  "Europe/N. Africa",
                           ifelse(meteorites_clean$reclat >= 25 & meteorites_clean$reclat <= 50 & 
                                  meteorites_clean$reclong >= -130 & meteorites_clean$reclong <= -60,
                                  "North America",
                           ifelse(meteorites_clean$reclat >= 15 & meteorites_clean$reclat <= 35 & 
                                  meteorites_clean$reclong >= -20 & meteorites_clean$reclong <= 55,
                                  "Sahara Desert",
                           ifelse(meteorites_clean$reclat >= -40 & meteorites_clean$reclat <= -10 & 
                                  meteorites_clean$reclong >= 110 & meteorites_clean$reclong <= 155,
                                  "Australia",
                                  "Other Regions")))))

# 8. Log transform mass for better visualization of wide range
meteorites_clean$log_mass <- log10(meteorites_clean$mass..g. + 1)
```

### Processing Summary

After data processing, the cleaned dataset contains `r nrow(meteorites_clean)` meteorite records spanning from `r min(meteorites_clean$year)` to `r max(meteorites_clean$year)`.

```{r summary-output, echo=FALSE, comment=''}
# Display simple summaries
cat("Discoveries by Time Period:\n")
print(table(meteorites_clean$time_period))

cat("\n\nDiscoveries by Meteorite Type:\n")
print(table(meteorites_clean$meteorite_type))

cat("\n\nDiscoveries by Region:\n")
print(table(meteorites_clean$region))
```

### Summary of Processing Steps:

1.  **Removed missing values**: Filtered out records with missing coordinates, years, or classification data to ensure complete data for analysis

2.  **Removed invalid coordinates**: Excluded entries with latitude/longitude outside valid geographic ranges (-90 to 90 for latitude, -180 to 180 for longitude)

3.  **Focused time period**: Limited analysis to 1800-2013 to focus on the modern era of meteorite science with better documentation

4.  **Created time period categories**: Grouped years into four meaningful periods (pre-1950, 1950-1979, 1980-1999, 2000-2013) to enable temporal comparison

5.  **Created meteorite type categories**: Simplified the detailed `recclass` variable into three major groups:

    -   **Stone meteorites**: Including all chondrites (H, L, LL, E, C types) and achondrites - the most common type
    -   **Iron meteorites**: Metallic meteorites composed primarily of iron-nickel alloys - heavy and distinctive
    -   **Stony-Iron meteorites**: Rare mixed composition meteorites (pallasites, mesosiderites)

6.  **Created geographic regions**: Classified meteorites into major discovery regions (Antarctica, Europe/N. Africa, North America, Sahara Desert, Australia) based on coordinate ranges

7.  **Log-transformed mass**: Applied log transformation to meteorite mass values to better visualize the wide range of sizes (from less than 1 gram to over 60,000 kg)

------------------------------------------------------------------------

## Problem 2: Reproduce the visualization from your "Data Story" assignment submission. Interpret the visualization and describe any key conclusions.

**Do different regions discover different types of meteorites, and has this changed over time?**

## Visualization

```{r main-visualization, fig.width=12, fig.height=10}
# Get world map data
world_map <- map_data("world")

# Define colors for meteorite types
type_colors <- c("Stone" = "#2A9D8F",      # Teal for Stone (most common)
                 "Iron" = "#E63946",        # Red for Iron (heavy, distinctive)
                 "Stony-Iron" = "#F4A261",  # Orange for Stony-Iron (rare)
                 "Other" = "#999999")       # Gray for Other

# Create faceted world map showing type patterns over time
ggplot() +
  # Base world map
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray70", size = 0.2) +
  # Meteorite landing points colored by type
  geom_point(data = meteorites_clean,
             aes(x = reclong, y = reclat, color = meteorite_type, size = log_mass),
             alpha = 0.6) +
  # Facet by time period to show temporal changes
  facet_wrap(~ time_period, ncol = 2) +
  # Color scale for meteorite types
  scale_color_manual(
    name = "Meteorite Type",
    values = type_colors
  ) +
  # Scale and labels
  scale_size_continuous(
    name = "Mass (g)",
    range = c(0.3, 3),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("10", "100", "1K", "10K", "100K")
  ) +
  labs(
    title = "Different Regions Discover Different Types of Meteorites Over Time",
    x = "Longitude",
    y = "Latitude",
    caption = "Data Source: NASA Open Data Portal / The Meteoritical Society (data.nasa.gov)\nNote: Point size represents meteorite mass (log scale); color represents composition type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray90", color = NA),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0)
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 4, alpha = 1), order = 1),
    size = guide_legend(order = 2)
  ) +
  coord_fixed(ratio = 1.3)
```

## Interpretation

The color-coded world maps reveal striking differences in the **types of meteorites** discovered across different regions and time periods:

### Early Period (1800-1949): Iron-Rich European Finds

The earliest time period shows meteorite discoveries in Europe and populated regions with a notably higher proportion of **Iron meteorites** (red points). This pattern reflects a strong discovery bias:

-   **Iron meteorites are easier to recognize**: Their metallic appearance makes them distinctive and obvious to non-specialists

-   **Better preservation**: Iron meteorites resist weathering much better than stone meteorites in temperate climates

-   **Heavier and more noticeable**: Iron meteorites have higher density, making them more likely to be picked up and identified

-   **Cultural recognition**: Metal objects from space were more likely to be considered noteworthy and preserved

### Mid-Century (1950-1979): Diversification in Desert Regions

As discoveries expanded to desert regions in North America, the Sahara, and Australia, we see a more balanced mix of meteorite types:

-   **Desert preservation**: Hot desert environments preserve both stone and iron meteorites well

-   **Visual contrast**: Both types are visible against light-colored sand

-   **Systematic searches begin**: Scientists start deliberately searching rather than relying on chance finds

-   **More representative sample**: The type distribution begins to reflect actual fall patterns

### Modern Era (1980-2013): Stone Meteorite Dominance in Antarctica

The Antarctic regions show an overwhelming predominance of **Stone meteorites** (teal points). This dramatic shift reveals:

-   **True fall distribution**: Stone meteorites comprise \~94% of observed meteorite falls, and Antarctica's preservation shows this reality

-   **Ice preservation**: Antarctica's cold, dry environment preserves fragile stone meteorites that would decompose elsewhere

-   **Systematic ice field searches**: Targeted search programs specifically look for all meteorite types

-   **Blue ice concentration**: Ice flow patterns naturally concentrate meteorites in searchable areas

### Geographic Composition Patterns

Clear regional differences emerge: - **Antarctica**: Predominantly stone meteorites (\~85-90%) - reflects true cosmic abundance

-   **Europe/North Africa (early periods)**: Higher iron proportion (30-40%) - preservation and recognition bias

-   **Sahara Desert**: Mixed - good preservation of both types

-   **North America**: Gradual shift from iron-rich to more balanced as systematic searches developed

## Key Conclusions

1.  **Discovery bias shapes our meteorite collection**: Early collections over-represent durable iron meteorites, while modern Antarctic surveys provide a more accurate picture of what actually falls from space.

2.  **Environment determines what survives to be found**:

    -   Temperate climates favor iron meteorite survival

    -   Desert environments (hot and cold) preserve both types

    -   Antarctica's ice provides the best preservation for fragile stone meteorites

3.  **Search methodology matters**: The evolution from chance finds in populated areas (favoring obvious iron meteorites) to systematic ice field surveys (capturing fragile stone meteorites) represents a fundamental shift in meteorite science toward obtaining truly representative samples.

4.  **Regional composition differences reveal preservation bias**: The changing type distribution across regions and time periods demonstrates that **where and how we search fundamentally shapes our understanding of extraterrestrial materials**.

This visualization powerfully illustrates how **environmental preservation, human observation capabilities, and scientific methodology** shape our extraterrestrial sample collection. The geographic and temporal patterns tell a compelling story about the evolution of planetary science.

```{r}
# ============================================================
# Graph 3: Stacked Bar Chart of Meteorite Composition by Era
# "What we're finding has shifted too"
# ============================================================

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Count meteorites by time period and type
composition_data <- meteorites_clean %>%
  group_by(time_period, meteorite_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(time_period) %>%
  mutate(
    total = sum(count),
    proportion = count / total * 100
  ) %>%
  ungroup()

# Define colors matching your original palette
type_colors <- c(
  "Stone"      = "#2A9D8F",
  "Iron"       = "#E63946",
  "Stony-Iron" = "#F4A261",
  "Other"      = "#999999"
)

# --- Build the Plot ---

ggplot(composition_data, 
       aes(x = time_period, y = proportion, fill = meteorite_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  # Percentage labels inside each segment (hide if too small)
  geom_text(
    aes(label = ifelse(proportion >= 4, 
                        paste0(round(proportion, 1), "%"), 
                        "")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 3.5
  ) +
  # Count annotations below each bar
  geom_text(
    data = composition_data %>%
      group_by(time_period) %>%
      summarise(total = first(total), .groups = "drop"),
    aes(x = time_period, y = -3, 
        label = paste0("n = ", format(total, big.mark = ",")),
        fill = NULL),
    size = 3,
    color = "gray50",
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    name = "Meteorite Type",
    values = type_colors
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.05, 0.02))
  ) +
  labs(
    title = "What We're Finding Has Shifted Too",
    subtitle = "Proportion of meteorite types discovered in each era",
    x = NULL,
    y = "Proportion of discoveries",
    caption = "Data Source: NASA Open Data Portal / The Meteoritical Society (data.nasa.gov)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0),
    plot.caption  = element_text(size = 8, color = "gray50", hjust = 0),
    axis.text.x   = element_text(size = 9, lineheight = 1.1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position = "bottom",
    legend.title    = element_text(face = "bold", size = 10),
    legend.text     = element_text(size = 9)
  ) +
  guides(fill = guide_legend(override.aes = list(size = 5)))
```

```{r}

# Graph 4: Violin / Box Plot of Mass by Era (Log-Scaled)
# "We're finding smaller meteorites now"


library(ggplot2)
library(dplyr)


# Remove rows with missing or zero mass for this plot
mass_data <- meteorites_clean %>%
  filter(!is.na(mass..g.) & mass..g. > 0)

# Define a palette for the four time periods
era_colors <- c(
  "1800-1949: Early Discoveries"  = "#E63946",
  "1950-1979: Mid-Century"        = "#F4A261",
  "1980-1999: Modern Era"         = "#2A9D8F",
  "2000-2013: Recent Discoveries" = "#457B9D"
)

# Compute summary statistics for annotation
mass_summary <- mass_data %>%
  group_by(time_period) %>%
  summarise(
    n        = n(),
    median_g = median(mass..g.),
    mean_g   = mean(mass..g.),
    .groups  = "drop"
  )

# Violin plot
ggplot(mass_data, aes(x = time_period, y = mass..g., fill = time_period)) +
  # Violin layer — shows full distribution shape
  geom_violin(
    alpha      = 0.35,
    color      = NA,
    scale      = "width",
    trim       = TRUE,
    adjust     = 0.8
  ) +
  # Box plot layer — shows quartiles and median
  geom_boxplot(
    width      = 0.15,
    outlier.shape = NA,
    color      = "gray30",
    fill       = "white",
    alpha      = 0.8
  ) +
  # Median annotation points
  stat_summary(
    fun    = median,
    geom   = "point",
    shape  = 18,
    size   = 3,
    color  = "gray20"
  ) +
  # Median value labels
  geom_text(
    data = mass_summary,
    aes(x = time_period, y = median_g,
        label = ifelse(median_g >= 1000,
                        paste0(round(median_g / 1000, 1), " kg"),
                        paste0(round(median_g), " g"))),
    hjust      = -0.3,
    vjust      = -0.8,
    size       = 3.2,
    fontface   = "bold",
    color      = "gray25",
    inherit.aes = FALSE
  ) +
  # Sample size labels at the bottom
  geom_text(
    data = mass_summary,
    aes(x = time_period, y = 0.05,
        label = paste0("n = ", format(n, big.mark = ","))),
    size       = 3,
    color      = "gray50",
    inherit.aes = FALSE
  ) +
  # Log scale for mass
  scale_y_log10(
    labels = function(x) {
      ifelse(x >= 1e6, paste0(x / 1e6, " t"),
      ifelse(x >= 1000, paste0(x / 1000, " kg"),
             paste0(x, " g")))
    },
    breaks = c(0.1, 1, 10, 100, 1000, 1e4, 1e5, 1e6, 1e7)
  ) +
  scale_fill_manual(values = era_colors) +
  scale_x_discrete(
    labels = c(
      "1800-1949: Early Discoveries"  = "1800-1949",
      "1950-1979: Mid-Century"        = "1950-1979",
      "1980-1999: Modern Era"         = "1980-1999",
      "2000-2013: Recent Discoveries" = "2000-2013"
    )
  ) +
  labs(
    title    = "We're Finding Smaller Meteorites Now",
    subtitle = "Mass distribution of discovered meteorites by era (log scale)\nMedian mass has dropped steadily as systematic searches detect smaller specimens",
    x        = NULL,
    y        = "Mass (log scale)",
    caption  = "Data Source: NASA Open Data Portal / The Meteoritical Society (data.nasa.gov)\nDiamond = median mass; box = IQR; \nn = number of meteorite discoveries per era; \noutliers suppressed for clarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle    = element_text(size = 10, color = "gray40", hjust = 0,
                                     lineheight = 1.2),
    plot.caption     = element_text(size = 8, color = "gray50", hjust = 0),
    axis.text.x      = element_text(size = 9, lineheight = 1.1),
    axis.text.y      = element_text(size = 9),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position  = "none"
  ) +
  # Reference line at 1 kg for intuitive anchor
  geom_hline(yintercept = 1000, linetype = "dashed", 
             color = "gray70", linewidth = 0.4) +
  annotate("text", x = 0.6, y = 1000, label = "1 kg",
           size = 2.8, color = "gray50", hjust = 0, vjust = -0.5)
```
